---
title: "S7 — Testing & Verification, Part 1: Unit Tests"
format: html
author: Florian Oswald
date: today
date-format: "D MMMM, YYYY"
---

```{r}
#| include: false
library(reticulate)
```

## Why this session exists

Since S2 you've been asking one question over and over: *"where is this wrong, and how do I check?"* So far the answer has mostly been "read the code, run it, look at the output." That doesn't scale, and it definitely doesn't scale once an agent is generating most of the code and you're reviewing rather than typing.

A **unit test** is a way of writing down "here is what I checked" so that it gets checked again automatically, every time the code changes — including every time an agent touches the code. This session is entirely remote/self-paced: work through it at your own computer, running everything as you go. S8 builds directly on this with property-based tests and golden-output checks; S7 is the foundation.

::: {.callout-tip}
## Objectives
- Know what a unit test is, and what it is *not* worth testing.
- Write tests in `testthat` (R) and `pytest` (Python) using arrange-act-assert.
- Understand fixtures: fixed, known inputs with a known correct output.
- Have a first test suite covering real data-wrangling transformations.
:::

## A stand-in for the S6 pipeline

S6 built a DuckDB pipeline on a real dataset. Whichever pipeline you built there, the testing habits below apply directly to it. To make this session self-contained, we'll test a small pipeline of the same shape here — Brexit referendum poll data (`data/brexit.csv`), read via DuckDB, with two transformations:

1. `check_shares()` — flags polls whose `remain + leave + undecided` doesn't sum to ~1. This is exactly the kind of silent data problem an agent will not flag on its own unless asked.
2. `pollster_summary()` — aggregates valid polls by pollster: number of polls, average remain/leave share.

The full pipeline code is in `scripts/s7/pipeline.R` and `scripts/s7/pipeline.py`. If you already have your own S6 pipeline, treat this as a worked example and then repeat the exercise on your own transformations.

::: {.panel-tabset}

### R

```{r}
source("scripts/s7/pipeline.R")
polls <- read_polls("data/brexit.csv")
checked <- check_shares(polls)
head(checked[, c("pollster", "remain", "leave", "undecided", "share_total", "valid")])
```

```{r}
pollster_summary(checked)
```

### Python

```{python}
import sys
sys.path.insert(0, "scripts/s7")
from pipeline import read_polls, check_shares, pollster_summary

polls = read_polls("data/brexit.csv")
checked = check_shares(polls)
checked[["pollster", "remain", "leave", "undecided", "share_total", "valid"]].head()
```

```{python}
pollster_summary(checked)
```

:::

Notice something already: some real polls in this dataset have shares that don't quite sum to 1 (rounding in the source data), and `check_shares()` has to decide on a tolerance for what counts as "close enough." That tolerance (`tol = 0.01` by default) is exactly the kind of decision that deserves a test — it's a number someone chose, and choices deserve to be pinned down.

## What to test (and what not to bother testing)

Rough triage, useful both when you write code by hand and when an agent writes a first draft for you to review:

**Worth testing:**

- Your own transformation logic — filtering rules, aggregation formulas, anything with a threshold or an edge case (empty input, all-invalid input, a single row).
- Anywhere a silent decision gets made (a tolerance, a default value, a "drop rows where X" rule an agent added without telling you).
- Anything that would be embarrassing or costly to get wrong quietly — a mean computed over the wrong denominator, a join that silently drops rows.

**Not worth testing:**

- Behaviour of a well-tested library you're calling correctly (don't re-test that `dplyr::filter()` filters, or that `pandas.DataFrame.mean()` computes a mean). Test *your usage* of it, not its internals.
- Trivial getters/setters or one-line wrappers with no logic.
- Anything that changes on every run for reasons outside your control (current date/time, live network calls) — unless you specifically design a test around that variability (e.g. mocking it out). We won't go there today; S8 covers that territory.

The line isn't always obvious, and that's fine — the question to ask is: *"if this were subtly wrong, would anything catch it before a reader of my output does?"* If the answer is no, write a test.

## Arrange — Act — Assert

Every test below (and every test you should write) has the same three-part shape:

1. **Arrange**: build the input. Ideally a small, fixed value you chose yourself — not real data pulled live, not whatever happens to be lying around.
2. **Act**: call the function under test, exactly once, with that input.
3. **Assert**: check the output against a value *you* computed by hand or reasoned out — never a value you got by running the code and copy-pasting whatever it printed. That verifies the code agrees with itself, not that it's correct.

::: {.panel-tabset}

### R (testthat)

```r
test_that("check_shares flags rows that sum to 1 as valid", {
  # Arrange
  polls <- make_polls()

  # Act
  out <- check_shares(polls)

  # Assert
  expect_equal(out$share_total, c(1.00, 0.98, 1.00, 1.00))
  expect_equal(out$valid, c(TRUE, FALSE, TRUE, TRUE))
})
```

### Python (pytest)

```python
def test_check_shares_flags_rows_that_sum_to_one(polls):
    # Arrange: `polls` fixture
    # Act
    out = check_shares(polls)
    # Assert
    assert list(out["share_total"]) == [1.00, 0.98, 1.00, 1.00]
    assert list(out["valid"]) == [True, False, True, True]
```

:::

Notice the `1.00, 0.98, 1.00, 1.00` and `[1.00, 0.98, 1.00, 1.00]` — those numbers were computed by hand from the fixture data below, not generated by running `check_shares()` and copying its output.

## Fixtures

A **fixture** is the "arrange" step, factored out so multiple tests can reuse it, and so it's obvious at a glance what the known input is. Both languages below build the same four-row table: two polls from pollster A (one valid, one that doesn't sum to 1), one valid poll from pollster B, and one poll from pollster C with an invalid (negative) sample size.

::: {.panel-tabset}

### R (a helper function)

```r
make_polls <- function() {
  data.frame(
    pollster   = c("A", "A", "B", "C"),
    samplesize = c(1000, 1000, 500, -1),
    remain     = c(0.52, 0.48, 0.50, 0.50),
    leave      = c(0.48, 0.50, 0.50, 0.50),
    undecided  = c(0.00, 0.00, 0.00, 0.00),
    stringsAsFactors = FALSE
  )
}
```

`testthat` has no special fixture decorator for this kind of case — a plain helper function called at the top of each `test_that()` block is idiomatic. (For fixtures that need setup/teardown, e.g. a temp file, see `withr::local_tempfile()` — out of scope today, but worth knowing the name.)

### Python (`@pytest.fixture`)

```python
@pytest.fixture
def polls() -> pd.DataFrame:
    return pd.DataFrame(
        {
            "pollster": ["A", "A", "B", "C"],
            "samplesize": [1000, 1000, 500, -1],
            "remain": [0.52, 0.48, 0.50, 0.50],
            "leave": [0.48, 0.50, 0.50, 0.50],
            "undecided": [0.00, 0.00, 0.00, 0.00],
        }
    )
```

`pytest` fixtures are functions decorated with `@pytest.fixture`; any test function that names `polls` as an argument receives a fresh copy automatically.

:::

## The full test suites

The complete, runnable test files:

- R: `scripts/s7/test-pipeline.R` (8 tests, run with `testthat::test_file("scripts/s7/test-pipeline.R")`)
- Python: `scripts/s7/test_pipeline.py` (6 tests, run with `pytest scripts/s7/test_pipeline.py -v`)

Both cover the same cases, because the pipelines are equivalent:

1. Shares that sum to 1 are flagged valid; shares that don't are flagged invalid.
2. A wider tolerance can let a borderline row through.
3. Invalid rows and rows with a non-positive sample size are dropped before aggregation.
4. Aggregation is numerically correct for a single surviving row.
5. **Edge case**: if *nothing* survives filtering, the summary is zero rows — not an error, not a row for the wrong pollster.
6. Missing required columns raise a clear error, rather than failing somewhere deep inside `dplyr`/`pandas` with a confusing message.

Case 5 is the one agents most often get wrong on a first draft: ask an agent to write `pollster_summary()` cold, and there's a good chance it doesn't handle the all-filtered-out case at all, or handles it inconsistently (an empty data frame with the wrong columns, or a `KeyError`/`NULL` crash). Writing that test *before* trusting the function is the whole point of this session.

Run them yourself:

```bash
Rscript -e 'testthat::test_file("scripts/s7/test-pipeline.R")'
```

```bash
pytest scripts/s7/test_pipeline.py -v
```

## Exercise

1. Open your own S6 pipeline (or, if you haven't finished S6, use `scripts/s7/pipeline.R` / `scripts/s7/pipeline.py` as-is).
2. Pick one transformation you have **not** tested yet. Ask your agent (OpenCode) to draft a `testthat` or `pytest` test file for it, using arrange-act-assert and at least one fixture.
3. Before running the tests, read the agent's draft and answer: does each assertion's expected value come from your own reasoning about the input, or did the agent just run the code and paste the output back as the "expected" value? The latter is not a test — it will always pass, and it will pass even after you introduce a bug.
4. Fix any assertions that fail the check in step 3, run the suite, and add one edge case the agent's draft missed (empty input is the classic one).
5. Note, in one or two sentences, what your agent's first draft got wrong or skipped. Bring it to S8 — this is exactly the kind of gap S8's expanded toolkit targets.

## Looking ahead

S8 goes beyond "does this specific input give this specific output" into invariants that must hold for *any* input (property-based testing), golden-output comparisons for outputs too complex to hand-compute, and assertions embedded directly in agent-facing scripts. The unit tests you wrote today are the floor, not the ceiling.
