---
title: "S2 — Programming Constructs, Agent-Native"
format: html
author: Florian Oswald
date: today
date-format: "D MMMM, YYYY"
---

```{r reticulate_config}
#| cache: false
#| include: false
library(reticulate)
```

::: {.callout-tip}
## Objectives
- Learn the four core building blocks of any language — **variable**, **loop**, **conditional**, **function** — in both R and Python.
- Practice the habit this whole course is built on: have an agent write a first draft, then **verify it, try to break it, and write a test that would have caught the bug if it were there.**
- By the end you should be asking, reflexively, of any code an agent hands you: *"where is this wrong, and how do I check?"*
:::

## Why we do this with an agent, not by hand

You could learn these four constructs the old way: read a syntax reference, type it in yourself, get red squiggly lines until it compiles. That's not how you'll actually write code anymore. You'll ask an agent — OpenCode, in this course — to draft something, and your job shifts from "get the syntax right" to "decide whether the result is *correct*."

Today we practice exactly that shift, on the simplest possible material, so the habit is cheap to build before the material gets harder in later sessions.

The core loop for the rest of this session, and honestly the rest of the course:

1. Ask the agent to write something.
2. Read what it produced. Does it look right?
3. Try to break it — pick an input the agent probably didn't think about.
4. Write a small check (an `assert` / `stopifnot`) that would catch the bug if it's there.
5. Run it. If it fails, you found something real. Fix it, and keep the check.

## Setup

Two terminals, one running `R`, one running `python` (a REPL, a script, or `opencode` sessions in each). We use `numpy` for arrays and `pandas` for dataframes in the reference section below.

```{python}
#| eval: false
import numpy as np
import pandas as pd
```

## Variables

A variable is a label attached to a value — a number, a string, or any more complex object.

::: {.panel-tabset group="language"}

### Python

```{python pvars}
#| cache: false
x = 12.3
x + 5
```

### R

```{r rvars}
#| cache: false
x <- 12.3  # `=` also works at top level, but `<-` is idiomatic
x + 5
```

:::

Where a variable is *visible* (its **scope**) matters. A function looks for names it doesn't recognize locally in the environment where it was *defined* (this is called **lexical scoping**), not necessarily where it's called from:

::: {.panel-tabset group="language"}

### Python

```{python pf}
#| cache: false
def myfun(y):
    return x + y  # `x` is not a parameter — where does it come from?

myfun(8)
```

### R

```{r rf}
#| cache: false
myfun <- function(y){
    x + y
}
myfun(8)
```

:::

Both work because `x` was defined earlier in the same scope the function was created in. If you'd never defined `x`, both would error — try it (comment out the `x <- ...` / `x = ...` lines above and re-run).

## Conditionals

A conditional runs a branch of code only when some condition is true. The shape is the same in every mainstream language: check a condition, do one thing if true, another (optionally) if false.

::: {.panel-tabset group="language"}

### Python

```{python pcond}
#| cache: false
def classify(n):
    if n > 0:
        return "positive"
    elif n < 0:
        return "negative"
    else:
        return "zero"

[classify(v) for v in [-3, 0, 4]]
```

### R

```{r rcond}
#| cache: false
classify <- function(n) {
  if (n > 0) {
    "positive"
  } else if (n < 0) {
    "negative"
  } else {
    "zero"
  }
}
sapply(c(-3, 0, 4), classify)
```

:::

Two things worth internalizing now, because they cause silent bugs later:

- **Comparing floats with `==` is dangerous.** `0.1 + 0.2 == 0.3` is `False` in Python and `FALSE` in R, because of floating-point representation. An agent will happily write `if (x == 0.3)` — it looks fine, it's wrong.
- **Empty/missing input often sails through a conditional without raising anything.** `if x > 0` where `x` is `None`/`NA` doesn't always fail loudly — sometimes it fails silently, or crashes far from where the real problem is. This is the kind of thing you have to go looking for; nothing forces you to notice it.

## Loops

Loops repeat an operation over a sequence, so you don't write near-identical statements by hand and don't have to touch the code again when the input size changes.

::: {.panel-tabset group="language"}

### Python

```{python ploop}
#| cache: false
for i in range(2, 5):
    print(f"this is number {i}")
```

### R

```{r rloop}
#| cache: false
for (i in 2:4) {
  print(paste("this is number", i))
}
```

:::

Look closely at what actually printed. If you intended to cover the numbers **2, 3, 4, 5**, did you get all four — and are the two bugs even the same bug? They're not, and that's the point:

- **Python's `range(2, 5)` is half-open by design**: it stops *before* its second argument, so it yields 2, 3, 4 and silently drops 5. This isn't a typo, it's how `range` always works — the bug is forgetting that convention when translating "2 through 5" into code.
- **R's `2:4` is not truncated by `:` at all** — R's `:` operator is inclusive on both ends, so `2:5` would correctly give 2, 3, 4, 5. The bug here is a plain wrong literal: whoever (or whatever agent) wrote `2:4` mistyped the upper bound, full stop.

Same visible symptom — one element missing — but two different root causes, in two different places (a language convention vs. a transcription slip). An agent draft can produce either kind, and "it printed one fewer number than expected" doesn't tell you which one you're looking at until you read the code.

::: {.callout-important}
## Exercise 1 — find the bug in each language, then check for it
1. Before reading further, predict what the two loops above print. Then run them.
2. Fix both so they print 2, 3, 4, **and** 5 — but explain in one sentence, per language, *why* the original was wrong (half-open range convention vs. wrong literal). If your explanation is the same for both, look again.
3. Write a check — `stopifnot()` in R, `assert` in Python — that captures the intended output (e.g. "the loop should touch exactly 4 numbers: 2,3,4,5") and would fail on the *original*, unfixed version. Save it; you'll use the same pattern below.
:::

## Functions

A function packages a computation so you can reuse it and give it a name that documents intent.

::: {.panel-tabset group="language"}

### Python

```{python pfun}
#| cache: false
def mean_positive(values):
    """Average of the positive numbers in `values`."""
    positives = [v for v in values if v > 0]
    return sum(positives) / len(positives)

mean_positive([1, -2, 3, 4])
```

### R

```{r rfun}
#| cache: false
mean_positive <- function(values) {
  positives <- values[values > 0]
  sum(positives) / length(positives)
}
mean_positive(c(1, -2, 3, 4))
```

:::

Run it on `[1, -2, 3, 4]` and it works fine: `2.6...`. Now think about what breaks it before you try it.

## The exercise: agent draft, then verify, then break it

This is the actual point of today. Below are two small tasks. For each, the "agent draft" columns show code exactly as a first draft from an agent might look — plausible, running, subtly wrong. Don't just read them: open OpenCode, ask it to write the same function fresh in both languages, and compare what you get against these. Free-tier models sometimes get these exactly right, sometimes reproduce the same bug — either way, work through the seeded version below so everyone in the room breaks the *same* code at least once.

### Task A — `mean_positive`: average of the positive numbers in a list

This is the function from above. It has a real bug: **what happens when there are no positive numbers?**

::: {.panel-tabset group="language"}

### Python (agent draft)

```{python pfun_bug}
#| cache: false
#| error: true
def mean_positive(values):
    positives = [v for v in values if v > 0]
    return sum(positives) / len(positives)

mean_positive([-1, -2, -3])
```

### R (agent draft)

```{r rfun_bug}
#| cache: false
#| error: true
mean_positive <- function(values) {
  positives <- values[values > 0]
  sum(positives) / length(positives)
}
mean_positive(c(-1, -2, -3))
```

:::

Python raises `ZeroDivisionError`. R silently returns `NaN` and moves on — no error at all, which is worse, because a silent `NaN` can propagate for pages of downstream code before anyone notices.

::: {.callout-important}
## Exercise 2 — write the test, then fix the function
1. Write a check that documents the bug before fixing anything:
   - Python: `assert len(positives) > 0, "no positive values in input"` (placed before the division).
   - R: `stopifnot(length(positives) > 0)`.
2. Confirm the check fails (raises) on `[-1, -2, -3]` / `c(-1, -2, -3)` in the *original* function.
3. Decide what the *correct* behavior should be — raise clearly? Return `NA`/`None` on purpose? There's no single right answer, but the agent's draft picked one **silently**, without telling you. Fix the function to do what you decided, on purpose, and keep the check as a permanent guard.
:::

### Task B — `running_total`: cumulative sum of a list

::: {.panel-tabset group="language"}

### Python (agent draft)

```{python prun_bug}
#| cache: false
def running_total(values):
    total = 0
    result = []
    for v in values[1:]:  # look closely at the slice
        total += v
        result.append(total)
    return result

running_total([10, 20, 30])
```

### R (agent draft)

```{r rrun_bug}
#| cache: false
running_total <- function(values) {
  total <- 0
  result <- c()
  for (v in values[-1]) {  # look closely at the indexing
    total <- total + v
    result <- c(result, total)
  }
  result
}
running_total(c(10, 20, 30))
```

:::

Both versions drop the first element (`values[1:]` in Python skips index 0; `values[-1]` in R excludes the first element) — so a 3-element input silently returns only 2 cumulative totals, and the first one is missing entirely from the running sum. This is the same *shape* of bug as Task A: it runs without error, produces output that looks like a plausible cumulative sum, and only close reading (or a test on length/first-element) catches it.

::: {.callout-important}
## Exercise 3 — same pattern, different bug
1. Predict, then check: what does `running_total([10, 20, 30])` return? What *should* it return? (Hint: length 3, first element 10.)
2. Write the test first — `assert len(running_total([10, 20, 30])) == 3` / `stopifnot(length(running_total(c(10,20,30))) == 3)` — confirm it fails against the draft.
3. Fix the function so the test passes. Also add an edge case: what should `running_total([])` return? Does your fix handle it? Write a check for that too.
:::

## Save your work

By the end of today you should have, in your S1 git repo (or a new folder within it):

- `constructs.py` and `constructs.R` — the corrected `mean_positive` and `running_total` functions in both languages.
- `test_constructs.py` (`assert` statements) and `test_constructs.R` (`stopifnot()` calls) — the checks you wrote that fail against the buggy drafts and pass against your fixes.

Commit them. We'll build directly on this repo structure in S3.

A worked reference solution (one way to fix both bugs, plus the checks that catch them) lives in [`scripts/s2-constructs/`](https://github.com/floswald/AgenticProgramming/tree/agentic-redesign/scripts/s2-constructs) in this course repo — don't look until you've written your own; the point is the exercise, not the answer key.

::: {.callout-note}
## A note on the tests above
The checks in this session are deliberately plain `assert` / `stopifnot()` — no test framework yet. That's on purpose: the point today is the *habit* of writing down what "correct" means and checking it, not test tooling. Frameworks (`pytest`, `testthat`) with fixtures, test discovery, and reporting come in S7, once you have a real pipeline worth testing systematically.
:::

## Reference: common data structures

You'll use these constantly; here's the vocabulary mapped between the two languages. (Not today's focus — a reference for when you need it.)

concept | Python | R
-----| -------|-----
1d list | `[1, 2]` | `c(1, 2)`
1d array | `np.array([1, 2])` | `c(1, 2)`
matrix | `np.array([[..],[..]])` | `matrix(data, rows, cols)`
n-d array | `np.array` (reshaped) | `array`
key-value map | `dict` | named `list`
dataframe | `pandas.DataFrame` | `data.frame`

::: {.panel-tabset group="language"}

### Python

```{python pref}
#| cache: false
import numpy as np
import pandas as pd

li = np.array([1, 3])
li * li          # element-by-element
d = {"peter": 1225, "alice": 4333}
pd.DataFrame({"one": [1.0, 2.0], "two": [4.0, 3.0]})
```

### R

```{r rref}
#| cache: false
li <- c(1, 3)
li * li          # element-by-element
d <- list(peter = 1225, alice = 4333)
data.frame(one = c(1, 2.0), two = c(4, 3.0))
```

:::
