---
title: "Data Wrangling: DuckDB"
format: clean-revealjs
author: Florian Oswald
subtitle: "[Agentic Programming](https://floswald.github.io/AgenticProgramming/)"
date: today
date-format: "D MMMM, YYYY"
---

```{r setup}
#| include: false
library(duckdb)
library(dplyr)
con <- dbConnect(duckdb())
polls <- tbl(con, "read_csv_auto('data/brexit.csv')")
```

## Where we are

* S4: how an LLM actually works (tokens in, tokens out, no ground truth inside).
* S5: how an *agent* wraps that model with tools, context, and a harness.
* Today the tool is a **database**, and the task is **data wrangling**: getting from raw files to trustworthy numbers.

::: {.callout-note}
## Today's habit, again
The agent will happily write a query that runs without error and returns a *plausible* number. Running without error is not the same as being right.
:::

---

## Today's plan

1. What DuckDB is, and why it is a good fit for this course.
2. One dataset, three ways to query it: **SQL**, **dplyr**, **Ibis**.
3. Let the agent draft a query. Find out it's wrong. Fix it.
4. You do the same, on your own query, and write it down.

# What is DuckDB?

---

## DuckDB in one slide

* An **in-process** analytical (OLAP) SQL database — like SQLite, but built for aggregations over large columnar data instead of many small transactions.
* No server to install, no daemon to start: `import duckdb` (Python), `library(duckdb)` (R), or the `duckdb` CLI binary.
* Reads CSV / Parquet / JSON **directly**, without a separate load step:

```sql
SELECT * FROM 'data/brexit.csv' LIMIT 5;
```

* Fast, columnar, and free — a real analytical engine on your laptop.

---

## Why DuckDB for *this* course

::: {.fragment}
* R and Python both talk SQL fluently through DuckDB — it's a genuine common ground between the two languages this course uses.
:::
::: {.fragment}
* The **same question** can be asked three ways:
  * raw **SQL** — the lingua franca, and what the agent tends to reach for first.
  * **dplyr** verbs (R) translated to SQL under the hood.
  * **Ibis** expressions (Python) translated to SQL under the hood.
:::
::: {.fragment}
* All three compile to the *same* DuckDB query plan. If they disagree, one of you (or the agent) made a mistake — that disagreement is itself a verification tool.
:::

---

## Setup

::: {.columns}
::: {.column width="33%"}
**SQL / CLI**
```bash
duckdb
```
:::
::: {.column width="33%"}
**R**
```r
library(duckdb)
library(dplyr)
con <- dbConnect(duckdb())
polls <- tbl(
  con,
  "read_csv_auto('data/brexit.csv')"
)
```
:::
::: {.column width="33%"}
**Python**
```python
import duckdb
import ibis

con = ibis.duckdb.connect()
polls = con.read_csv(
  "data/brexit.csv",
  table_name="polls"
)
```
:::
:::

All three point at the *same file* — no import/export, no separate database to keep in sync.

# The dataset

---

## UK EU-referendum polls (2016)

* `data/brexit.csv`: 127 opinion polls conducted between January and June 2016, ahead of the UK's "Brexit" referendum.
* Columns: `pollster`, `poll_type`, `startdate`/`enddate`, `samplesize`, `remain`, `leave`, `undecided`, `spread`.
* **Known answer**: the actual referendum result was **Leave 51.89%, Remain 48.11%** — a fixed, external fact we can check any query's output against.

```{sql, connection=con}
SELECT pollster, enddate, samplesize, remain, leave, undecided
FROM 'data/brexit.csv'
ORDER BY enddate DESC
LIMIT 5;
```

---

## Ask the agent: "what does this data predict?"

Prompt given to the agent (OpenCode):

> *"Write a DuckDB SQL query against `data/brexit.csv` that estimates the referendum outcome — the percentage voting Leave and the percentage voting Remain."*

The agent's first draft — plausible, runs without error:

```sql
SELECT
  round(avg(leave)  * 100, 2) AS leave_pct,
  round(avg(remain) * 100, 2) AS remain_pct
FROM 'data/brexit.csv';
```

---

## Running the agent's first query

```{sql, connection=con}
SELECT
  round(avg(leave)  * 100, 2) AS leave_pct,
  round(avg(remain) * 100, 2) AS remain_pct
FROM 'data/brexit.csv';
```

::: {.fragment}
`leave_pct = 42.23`, `remain_pct = 44.24`. Looks like a plausible pair of vote shares. **Do not trust it yet.**
:::

---

## Verify before you trust it

The cheapest check costs one extra line: do the shares **add up to 100%**?

```{sql, connection=con}
SELECT
  round(avg(leave)  * 100, 2) AS leave_pct,
  round(avg(remain) * 100, 2) AS remain_pct,
  round(avg(undecided) * 100, 2) AS undecided_pct,
  round(avg(leave + remain + undecided) * 100, 2) AS total_pct
FROM 'data/brexit.csv';
```

::: {.fragment}
They do (≈100%, modulo rounding) — but the query **silently dropped `undecided` respondents** from the two numbers it reported. `leave_pct + remain_pct` alone is only 86.5%, not 100%. The agent answered the wrong question: "average raw share" instead of "share of those who expressed a preference."
:::

---

## What actually went wrong

::: {.fragment}
1. **No re-normalization**: it reported `leave` and `remain` shares out of *all* respondents (including undecided), not as shares of decided voters — the quantity we actually asked for.
:::
::: {.fragment}
2. **No weighting**: `AVG()` treats a poll of 500 people the same as a poll of 5,000 — every poll gets an equal vote, regardless of `samplesize`.
:::
::: {.fragment}
3. **No date filtering**: it averaged polls from January *and* June together, as if opinion in January (5+ months before the vote) were as informative as opinion the week before.
:::

None of these threw an error. All three are silent, plausible-looking mistakes — exactly the kind unit tests and invariant checks (S7–S8) are built to catch.

---

## Fixed query — SQL

```{sql, connection=con}
SELECT
  round(sum(leave  / (leave + remain) * samplesize) / sum(samplesize) * 100, 2) AS leave_pct,
  round(sum(remain / (leave + remain) * samplesize) / sum(samplesize) * 100, 2) AS remain_pct,
  count(*) AS n_polls
FROM 'data/brexit.csv'
WHERE enddate >= DATE '2016-06-16';   -- final week before the vote
```

Re-normalized (excludes undecided from the denominator), sample-size weighted, restricted to the final week.

---

## Same query — dplyr (R)

```{r}
weighted_r <- polls |>
  filter(enddate >= as.Date("2016-06-16")) |>
  summarise(
    leave_pct  = round(sum(leave  / (leave + remain) * samplesize) / sum(samplesize) * 100, 2),
    remain_pct = round(sum(remain / (leave + remain) * samplesize) / sum(samplesize) * 100, 2),
    n_polls    = n()
  )
weighted_r
```

`tbl(con, ...)` on a DuckDB connection means these dplyr verbs are translated to SQL and run *inside* DuckDB — `dplyr::show_query(weighted_r)` prints the generated SQL, useful for checking the agent's dplyr code actually does what you think.

---

## Same query — Ibis (Python)

```{python}
import ibis

con2 = ibis.duckdb.connect()
polls_py = con2.read_csv("data/brexit.csv", table_name="polls")

recent = polls_py.filter(polls_py.enddate >= ibis.literal("2016-06-16").cast("date"))
w = recent.samplesize
share_leave  = (recent.leave  / (recent.leave + recent.remain)) * w
share_remain = (recent.remain / (recent.leave + recent.remain)) * w

weighted_py = recent.aggregate(
    leave_pct=((share_leave.sum()  / w.sum()) * 100).round(2),
    remain_pct=((share_remain.sum() / w.sum()) * 100).round(2),
    n_polls=recent.count(),
)
weighted_py.execute()
```

Same numbers as the SQL and the dplyr version — three languages, one DuckDB query plan.

---

## Check against the known answer

* Fixed query, final week of polling: **Leave ≈ 48.6%, Remain ≈ 51.4%**.
* Actual result: **Leave 51.89%, Remain 48.11%**.

::: {.callout-warning}
## The query is now correct. The prediction is still wrong.
Fixing the SQL bug (undecided handling, weighting, date window) got us a defensible number — but it's still on the wrong side of the actual outcome. 2016 UK polls famously under-counted Leave support (non-response bias, "shy" Leave voters, herding among pollsters).
:::

This is the distinction to keep: **verification catches query bugs** (wrong denominator, wrong weights, wrong date range — things you can check against an invariant). It does **not** catch **measurement error in the data itself** — that needs domain knowledge, not a better query. Keep this distinction for S9 (cost/quality tradeoffs) and S11 (cataloguing failure modes).

# Your turn

---

## Exercise (20 min)

Working with the same `brexit.csv` (or a dataset of your choice with a known, checkable answer):

1. Give your agent a data question and let it draft a **first** query — SQL, dplyr, or Ibis, your choice.
2. Run it. Get a number.
3. **Before you believe it**, write down at least one check: does it sum to what it should? Does it match a known external fact? Does a second language (SQL vs. dplyr vs. Ibis) give the same answer?
4. If the first answer fails a check, find out why, fix the query, and re-run.

Ideas to try: split by `pollster` or `poll_type` (phone vs. online — did they differ systematically?), compute a rolling weekly average, or compare `spread` in the data to `leave - remain` you compute yourself (does the pollster's number match your recomputation?).

---

## Deliverable: the written note

For the session outcome, write down **one concrete case** where the agent's first answer was wrong, in this format:

::: {.callout-tip}
## Template
- **Prompt** you gave the agent.
- **First query** it produced (SQL/dplyr/Ibis).
- **First output** — the specific wrong number(s).
- **How you caught it** — the check that failed (invariant, known answer, cross-language comparison, or something else).
- **Fixed query** and the corrected output.
:::

This is not a hypothetical exercise — use whatever actually happened when you ran your own prompt. If your agent got it right first try, push it further (harder aggregation, trickier date logic, a join) until it doesn't.

---

## Wrap-up

* DuckDB gives you one query engine reachable from SQL, R (dplyr), and Python (Ibis) — no data export/import between them.
* The agent's first query is a draft, not an answer. It ran without error and still ignored undecided respondents, sample weights, and the relevant date window.
* Checking against an invariant (do shares sum to 100%?) or a known external fact (the actual referendum result) is cheap and catches real bugs.
* Even a *correct* query only answers the question you asked — it can't fix bias in the underlying data.
* Next (S7): turning these ad-hoc checks into a proper, repeatable **test suite**.

```{r cleanup}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```
